Quantifying the impacts of bias in landcover data on global change analyses

### 2015-09-11 paper/main.tex
Finished draft through results. Updated Figure 6 (agent-bias), updated ET bias map, added Appendix to repo

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}

@@

We undertook five further analyses to investigate how map error can impact assessments founded on landcover maps. These include first-order analyses, in which values for a variable of interest are mapped to particular cover type(s), and second-order analyses, in which a process model draws on the cover types' values to calculate an output value. We created four datasets to represent second order analyses. The first was a series of maps of vegetated carbon stocks created following the methodology of Ruesch and Gibbs' \cite{ruesch_new_2008}. The second was constrained cropland percentage maps, which, following Ramankutty et al \cite{ramankutty_farming_2008} were adjusted so that their total cropland areas matched provincial-level reported cropland totals. Using these adjusted cropland percentage maps, we disaggregated district-reported maize harvested area and yields \cite[following][]{monfreda_farming_2008} . We then compared differences between total carbon stock estimates calculated from the reference map with those from the four cropland maps, and again examined how these differences changed as a function of aggregation scale.  We made the same comparisons for total maize harvested area, average yield, and total production.

@@

Absolute mean biases in yield were also substantial, and generally 10-15\% larger than production biases across all spatial scales, except for GlobCover where absolute production biases exceeded yield bias at 5-100 km of aggregation .

\subsection{Bias in evapotranspiration estimates}
\begin{itemize}
 \item Spatial biases most pronounced for VIC run with MODIS and GlobCover. These two overestimate ET along the southern margin, but typically by less than 5-10\%. (Fig. 5).  In the interior, VIC-GlobCover switches between under-  and over-estimation and then back again moving from West to East, but the bias is on the order of 5\% magnitude.  MODIS shows a similar pattern.    
 \item GeoWiki and SA-LC shows much less discrepancy in interior. Greatest tendency is to underestimate along the southern coast. 
 \item Difference in country-level ET estimates negligible ($<$2\%).
 \item \textcolor{red}{\textbf{Justin}, can you give some thoughts on reasons for these patterns? The tendency seems to be that where cropland is underestimated (blue areas in Fig. 1) then ET is overestimated (red areas in Fig. 5). Here I calculate all difference as (ref - lc) / ref * 100, with lc being one of the four landcover maps. My expectation was that crop ET was higher than most surrounding vegetation types, so one would expect ET to be underestimated where cropland is underestimated?}
\end{itemize}

\vspace{-0.5 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/et_bias_map.pdf}}
\caption{Differences in evapotranpiration estimates from a 29 year run of the VIC land surface hydrology model when initialized with LAI response curves derived from the reference map, versus those from the four cropland maps. Columns indicate differences between ET estimates from different portions of the model's time series: annual mean ET; highest annual ET value in the time series (TS max); smallest annual ET value in the time series (TS min); total of mean monthly ET for the three months having the highest ET during the year (Peak 3); the mean monthly ET of the peak ET month.}\label{afoto}
\end{figure}

\subsection{Initialization errors in spatial agent-based models}

\begin{itemize}
  \item We selected four magisterial districts with similar climates, and differing abundances of cropland ranging from X-Y\%, and differing patterns of map bias (Fig. S8). 
  \item Using household survey statistics from a project in Zambia that provides information on the average field area of farming households (X ha, SD = Y ha), we assigned each district a number of households determined by the total area of reference cropland in each district divided by mean farm size. \textcolor{red}{\textbf{Tom, Peng}, please correct as needed--maybe there is some variability to account for agent heterogeneity drawn by resampling from the distribution?}. This number of households represents the ``true'' number of households in the district.
  \item Disaggregated cropland to 100 m resolution, which is required by the model. 
  \item An agent-based model was then initialized wherein the household locations were seeded throughout the district, and assigned the required amount of cropland with a radius of X from the household location.
  \item These steps were repeated for each cropland map using the same number of household agents. 
  \item We then examined the differences that occurred between the agent initialization from the reference map and each of the four cropland maps, in terms of the percent of households that did not have cropland allocated, and the amount of cropland left unallocated in each district (Fig. 6).  
  \item The percent households without farms is 0 when cropland is overestimated (because there is plenty of land to choose from), but increases in a 1:1 relationship with the amount of cropland underestimation (positive values on X-axis in Fig. 6). \end{itemize}

\vspace{-0.5 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.3\textwidth]{figures/agent-bias.pdf}}
\caption{Biases in agent-based model initialization relative to cropland map biases, measured in terms of the percent of households having no cropland allocated (top) and the percent of cropland left unallocated (bottom). Dot sizes represent a district number (smallest = district 1; largest = district 4), colors represent the landcover map.  \textcolor{red}{to fix: add x-axis legend, add district number to legend.}}
\label{afoto}
\end{figure}

\begin{itemize}
  \item The percent of land unallocated, however, shows a U-shaped relationship with cropland biases, declining to near zero when cropland bias is between zero and up to $\sim$50\% under-estimated, thereafter increasing. Thus in cases where cropland bias insufficient farmland for households, a signficant share of available cropland can go unallocated (nearly 15\%).  \textcolor{red}{\textbf{Tom, Peng}, I assume this has to do something with seeding and the spatial pattern of cropland, but not sure exactly what. Model not sure which HH to give the land too because there are too many for the amount of cropland available, so it gives up? }
\end{itemize}

%large gradient ($\sim \displaystyle{ \inlinefrac{1}{\delta}}$).

>>>>

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
\usepackage{placeins}

@@

We undertook five further analyses to investigate how map error can impact assessments founded on landcover maps. These include first-order analyses, in which values for a variable of interest are mapped to particular cover type(s), and second-order analyses, in which a process model draws on the cover types' values to calculate an output value. We created four datasets to represent first order analyses. The first was a series of maps of vegetated carbon stocks created following the methodology of Ruesch and Gibbs' \cite{ruesch_new_2008}. The second was constrained cropland percentage maps, which, following Ramankutty et al \cite{ramankutty_farming_2008} were adjusted so that their total cropland areas matched provincial-level reported cropland totals. Using these adjusted cropland percentage maps, we disaggregated district-reported maize harvested area and yields \cite[following][]{monfreda_farming_2008} . We then compared differences between total carbon stock estimates calculated from the reference map with those from the four cropland maps, and again examined how these differences changed as a function of aggregation scale.  We made the same comparisons for total maize harvested area, average yield, and total production.

@@

Absolute mean biases in yield were also substantial, and generally 10-15\% larger than production biases across all spatial scales, except for GlobCover where absolute production biases exceeded yield bias at 5-100 km of aggregation.

\subsection{Bias in evapotranspiration estimates}
Compared to carbon and crop related examples, actual and absolute biases in evapotranspiration calculated using the VIC model were small and averaged to less than +/-1\%.  There are several hotspots of discrepancy, however, which are evident in maps of spatial biases (Fig. 6). The most pronounced of these are the 5-15\% overestimates by VIC initialized with MODIS and GlobCover in the center of the country, and along the southern and western coasts, particularly the latter where the ET overestimates reach 25\%. These locations correspond primarily to margins of major crop production regions--in the center is the westernmost boundary of the summer rainfall growing region, where maize is the primary crop, while the bias hotspot up the west coast marks one of the two boundaries of the winter rainfall region where wheat farming dominates\cite{hardy_rainfed_2011}. These two boundaries correspond respectively to approximately 400 and 200 mm of growing season rainfall.

SA-LC and GeoWiki also produced biased ET estimates along the southern and western coasts, but here the tendency was to underestimate ET, while biases in the center of the country were either negligible to absent.  All but MODIS underestimated ET by 5-15\% in the northern tip of the country.

\FloatBarrier
\begin{figure}[!ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/et_bias_map.pdf}}
\caption{Differences in annual mean evapotranspiration estimates from 29-year runs of the VIC land surface hydrology model when initialized with LAI response curves derived from the reference map, versus those from the four cropland maps.}\label{afoto}
\end{figure}

\subsection{Initialization biases in agent-based models}
We used an agent-based model of food security that represents the interactions between hundreds of individual farming households over multiple seasons \cite{chen_dependency_2013}. The model is thus computationally intensive, and, like many spatial ABMs, run over smaller geographic domains (e.g. districts, rather than an entire country) and at higher spatial resolutions (10s to 100s of meters) that are needed to represent the different land units of single farmers. To match these computational characteristics, we selected four contiguous magisterial districts (ranging from 1,040-1,343 km$^2$, Fig. S8) in the eastern part of the country, having between 28-45\% of their areas devoted to cropland, according to the reference map.

We disaggregated the cropland percentages in all maps to binary cropland/non-cropland cover types with 100 m resolution, which matches the typical field size (1 ha) for smallholder farmers in household survey data (collected in Zambia) used in developing the agent-based model \cite{chen_dependency_2013}. These surveys found mean household crop field area to be 2 ha, which we divided into reference cropland areas to estimate the total households within each district. We then initialized the model by assigning each household agent two cropland pixels. In order to emulate the natural groupings of communities, the model only assigns a household fields that are within 1.5 km of other agents' fields, and if those pixels have not already been allocated to another agent. The model thus iteratively grows ``communities'' until all households are assigned cropland, or all available cropland is allocated.

We used each of the five maps to initialize the model, and assessed how bias in cropland cover estimates impacted the initialization process, in terms of how many agents were not assigned fields, and how many fields were left unallocated. For the first metric, there was a one-to-one relationship between the percent by which cropland area underestimated and the percent of households left without and farmland (Fig. 6, left panel), the most extreme examples occurring when MODIS cropland was used to initialize the model in districts 1 and 2, where $\sim$85\% of agents could not be allocated farmland because MODIS underestimated cropland by that amount $\sim$85\%. No agents were left without fields where maps overestimated cropland extent (GeoWiki, SA-LC), but in these cases the percent of cropland left unallocated matched the size of the overestimate (e.g. $\sim$20\% for SA-LC, Fig. 6 right panel). Interestingly, the overall relationship between land allocation and cropland bias was U-shaped, as the model also failed to give land to agents when cropland was underestimated by 50\% or more (Fig. 6, right panel), as exemplified by the MODIS result for districts 1 and 2, where 7-12\% of cropland was left unallocated despite the fact that only 85\% of agents had no land.  This curious relationship occurred because cropland tends to cluster, and when it is underestimated, the size of these clusters is small, resulting in islands of cropland that fall outside of the search radius (which is constrained by an absolute distance and the proximity of other agents) within which cropland is sought when agents are seeded onto the landscape.

%randomly siting 100 household agents within each district, and allocating the nearest two cropland pixels to each household. The remaining agents are then iteratively assigned unallocated cropland pixels within a 1.5 km radius of existing agents' fields, and this process continues until all agents are assigned cropland, or all available cropland is allocated. This initialization process

\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/agent-bias.pdf}}
\caption{Biases in agent-based model initialization relative to cropland map biases, measured in terms of the percent of households having no cropland allocated (left) and the percent of cropland left unallocated (right). Dot sizes correspond to district numbers, colors represent the landcover map.}
\label{afoto}
\end{figure}
